---
title: "Leucegene Validation"
output:
  workflowr::wflow_html:
      code_folding: hide
---

```{r knitr, include = FALSE}
DOCNAME = "Leucegene_Validation"
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 9,
                      fig.height     = 5,
                      message        = FALSE,
                      warning        = FALSE)
```


```{r libraries, cache = FALSE}
# util
library(data.table)
library(dplyr)
library(here)
library(stringr)

# plotting/tables
library(ggplot2)
library(gt)
```


```{r options, cache = FALSE}
options(stringsAsFactors = FALSE)
```


```{r source, cache = FALSE}
source(here("code/leucegene_helper.R"))
```

Leucegene Validation
====================

Here we analyse the results of MINTIE run on a number of Leucegene samples,
including the effect of controls on a cohort with KMT2A-PTD variants. We also
check whether MINTIE has called known variants within the cohort.

```{r load_data, cache = FALSE}
# load SRX to patient ID lookup table
kmt2a_patient_lookup <- read.delim(here("data/leucegene/sample_info/KMT2A-PTD_samples.txt"),
                                   header = FALSE,
                                   col.names = c("sample", "patient"))

kmt2a_results_dir <- here("data/leucegene/KMT2A-PTD_results")

# load KMT2A cohort comparisons against all other controls
kmt2a_results <- load_controls_comparison(kmt2a_results_dir)
kmt2a_results <- inner_join(kmt2a_results, kmt2a_patient_lookup, by = "sample")

# load other validation reults and truth table
truth <- read.delim(here("data/leucegene/sample_info/variant_validation_table.tsv"), sep = "\t")
leucegene_results_dir <- here("data/leucegene/validation_results/MINTIE/")
validation <- list.files(leucegene_results_dir, full.names = TRUE) %>%
                lapply(., read.delim) %>%
                rbindlist(fill = TRUE) %>%
                filter(logFC > 5)
```

## KMT2A-PTD controls comparison

MINTIE paper Supplementary Figure 2. Shows the number of variant genes found in the Leucegene
cohort containing KMT2A PTDs. 

```{r KMT2A-PTD_controls_comparison, fig.height = 10, fig.width = 8}
results_summary <- get_results_summary(mutate(kmt2a_results,
                                              sample = patient,
                                              group_var = controls),
                                       group_var_name = "controls")

# build table
results_summary %>%
    group_by(controls) %>%
    summarise(min = min(V1), median = median(V1), max = max(V1)) %>%
    data.frame() %>%
    gt() %>%
     tab_header(
        title = md("**Total variant genes called using different controls**")
    ) %>%
    tab_options(
        table.font.size = 12
    ) %>%
    cols_label(
        controls = md("**Controls**"),
        min = md("**Min**"),
        median = md("**Median**"),
        max = md("**Max**")
    ) 

ggplot(results_summary, aes(sample, V1, fill = controls)) +
    geom_bar(position = position_dodge2(width = 0.9, preserve = "single"), stat = "identity") +
    theme_bw() +
    xlab("") +
    ylab("Genes with variants") +
    coord_flip() +
    theme(legend.position = "bottom") +
    scale_fill_brewer(palette="Dark2",
                      labels =  c("AML_controls" = "13 AMLs",
                                  "normal_controls" = "13 normals",
                                  "normal_controls_reduced" = "3 normals",
                                  "no_controls" = "No controls"))
```

## KMT2A variants found in cohort

MINTIE paper Supplementary Table 1. Shows whether MINITE found a KMT2A SV in each sample
for the given control group. Coverage obtained from [Audemard et al](https://doi.org/10.1101/295808).
[spreadsheet containing the Leucegene results](https://www.biorxiv.org/highwire/filestream/93175/field_highwire_adjunct_files/1/295808-2.zip)
must be manually added to `data/leucegene/sample_info` to run the code.

```{r KMT2A-PTDs_in_cohort}
# load results from km paper for coverage of KMT2A PTDs
kmt2a_lgene_km_results <- read.csv(here("data/leucegene/sample_info/KMT2A-PTD_8-2.fa.xls"), sep="\t") %>%
                            mutate(patient = Sample) %>%
                            group_by(patient) %>%
                            summarise(coverage = max(Min.coverage))

# check whether MINTIE found a KMT2A SV in each control set
found_using_cancon <- get_samples_with_kmt2a_sv(kmt2a_results, "AML_controls")
found_using_norcon <- get_samples_with_kmt2a_sv(kmt2a_results, "normal_controls")
found_using_redcon <- get_samples_with_kmt2a_sv(kmt2a_results, "normal_controls_reduced")
found_using_nocon  <- get_samples_with_kmt2a_sv(kmt2a_results, "no_controls")

# make the table
kmt2a_control_comp <- inner_join(kmt2a_patient_lookup, kmt2a_lgene_km_results, by = "patient") %>%
                        arrange(desc(coverage))
kmt2a_control_comp$`13_AMLs` <- kmt2a_control_comp$sample %in% found_using_cancon
kmt2a_control_comp$`13_normals` <- kmt2a_control_comp$sample %in% found_using_norcon
kmt2a_control_comp$`3_normals` <- kmt2a_control_comp$sample %in% found_using_redcon
kmt2a_control_comp$`no_controls` <- kmt2a_control_comp$sample %in% found_using_nocon

# build output table
kmt2a_control_comp %>%
    gt() %>% 
    cols_label(
        sample = md("**Sample**"),
        patient = md("**Patient**"),
        coverage = md("**Coverage**"),
        `13_AMLs` = md("**13 AMLs**"),
        `13_normals` = md("**13 Normals**"),
        `3_normals` = md("**3 Normals**"),
        `no_controls` = md("**No Controls**")
    ) %>%
    tab_header(
        title = md("**KMT2A PTDs found in Leucegene cohort**")
    ) %>%
    tab_options(
        table.font.size = 12
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(`13_AMLs`),
            rows = `13_AMLs`)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(`13_normals`),
            rows = `13_normals`)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(`3_normals`),
            rows = `3_normals`)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(`no_controls`),
            rows = `no_controls`)
    )
```

## Leucegene variants found by MINTIE

```{r leucegene_variants_called}
# add KMT2A results against AML controls as validation
validation <- filter(kmt2a_results, controls == "normal_controls") %>%
                select(-c(controls, patient)) %>%
                select(colnames(validation)) %>%
                rbind(., validation)

get_results_by_gene(validation) %>%
    group_by(sample) %>%
    summarise(vargenes = length(unique(gene))) %>%
    summarise(min = min(vargenes),
              median = median(vargenes),
              max = max(vargenes)) %>%
    data.frame() %>%
    gt() %>%
    tab_header(
        title = md("**Total MINTIE variant genes called by sample**")
    ) %>%
    tab_options(
        table.font.size = 12
    ) %>%
    cols_label(
        min = md("**Min**"),
        median = md("**Median**"),
        max = md("**Max**")
    )
```

### Summary of variants found by MINTIE, Arriba, Squid and TAP

Note that the TAP results must be obtained from Supplementary Table 4 from Chiu et al. 2018.

```{r leucegene_variants_validation}
# load other callers
arriba_results <- get_results(here("data/leucegene/validation_results/Arriba"))
squid_results <- get_results(here("data/leucegene/validation_results/Squid"))
tap_results <- read.delim(here("data/leucegene/validation_results/TAP/TAP_leucegene_results.txt"), sep = "\t")

# get variant gene locs (needed to check Squid results)
vargene_locs <- read.delim(here("data/leucegene/leucegene_vargene_locs.tsv"), sep = "\t")
vgx <- GRanges(seqnames = vargene_locs$chrom,
               ranges = IRanges(start = vargene_locs$start, end = vargene_locs$end),
               genes = vargene_locs$gene)

# make truth table
truth_table <- rowwise(truth) %>% 
    mutate(mintie_found = is_variant_in_mintie_results(Experiment, gene1, gene2, variant, validation)) %>%
    mutate(arriba_found = is_variant_in_results(Experiment, gene1, gene2, variant, "arriba", arriba_results)) %>%
    mutate(tap_found = is_variant_in_results(patient_ID, gene1, gene2, variant, "tap", tap_results)) %>%
    mutate(squid_found = is_variant_in_squid_results(Experiment, gene1, gene2, vgx, squid_results)) %>%
    data.frame()

gt(truth_table) %>%
     tab_header(
        title = md("**Variants found in Leucegene cohort**")
    ) %>%
    cols_label(
        patient_ID = md("**Patient**"),
        Experiment = md("**Experiment**"),
        gene1 = md("**Gene 1**"),
        gene2 = md("**Gene 2**"),
        variant = md("**Variant**"),
        cohort = md("**Cohort**"),
        mintie_found = md("**MINTIE Found**"),
        arriba_found = md("**Arriba Found**"),
        squid_found = md("**Squid Found**"),
        tap_found = md("**TAP Found**")
    ) %>%
    tab_options(
        table.font.size = 12
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(mintie_found),
            rows = mintie_found)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(arriba_found),
            rows = arriba_found)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(squid_found),
            rows = squid_found)
    ) %>%
    tab_style(
        style = cell_fill(color = "lightgreen"),
        locations = cells_body(
            columns = vars(tap_found),
            rows = tap_found)
    )

truth_table %>%
    group_by(gene1, gene2, variant) %>%
    summarise(mintie_detected = sum(mintie_found),
              arriba_detected = sum(arriba_found),
              squid_detected = sum(squid_found),
              tap_detected = sum(tap_found),
              total = length(mintie_found)) %>%
    data.table() %>%
    gt() %>%
    tab_header(
        title = md("**Summary of variants found in Leucegene cohort**")
    ) %>%
    cols_label(
        gene1 = md("**Gene 1**"),
        gene2 = md("**Gene 2**"),
        variant = md("**Variant**"),
        mintie_detected = md("**MINTIE Detected**"),
        arriba_detected = md("**Arriba Detected**"),
        squid_detected = md("**Squid Detected**"),
        tap_detected = md("**TAP Detected**"),
        total = md("**Total**")
    ) %>%
    tab_options(
        table.font.size = 12
    )
```



