---
title: "RCH B-ALL"
output:
  workflowr::wflow_html:
      code_folding: hide
---

```{r knitr, include = FALSE}
DOCNAME = "RCH_B-ALL"
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 9,
                      fig.height     = 5,
                      message        = FALSE,
                      warning        = FALSE)
```


```{r libraries, cache = FALSE}
# util
library(data.table)
library(dplyr)
library(here)
library(stringr)

# plotting/tables
library(ggplot2)
library(gt)
```


```{r options, cache = FALSE}
options(stringsAsFactors = FALSE)
```


```{r source, cache = FALSE}
source(here("code/leucegene_helper.R"))
```


RCH B-ALL results
=================

Here we analyse the results of MINTIE run on the RCH B-ALL cohort.

```{r load_data, cache = FALSE}
rch_ball_results_dir <- here("data/RCH_B-ALL")
rch_ball_results <- list.files(rch_ball_results_dir, full.names = TRUE) %>%
                        lapply(., read.delim) %>%
                        rbindlist() %>%
                        filter(logFC > 5)

# rename IDs to be consistent with doi: 10.1182/bloodadvances.2019001008
rch_ball_results$sample <- rch_ball_results$sample %>% 
                            str_split("^EKL-|^EKL|^PE15R-MLM-") %>%
                            lapply(., str_c, collapse = "") %>%
                            unlist() %>%
                            str_c("B-ALL_", .)

# list of ALL-associated genes
all_genes <- read.delim(here("data/ref/ALL_associated_genes.txt"), header=FALSE)$V1
```

## Variant genes called

Supplementary Figure 4 in the MINTIE paper. Shows the overall number of variant
genes called by MINTIE in the RCH B-ALL cohort.

```{r results_overview}
results_by_gene <- get_results_by_gene(rch_ball_results)
results_summary <- results_by_gene[, length(unique(gene)), by = "sample"]
results_summary <- results_summary %>% arrange(desc(V1))
results_summary$sample <- factor(results_summary$sample,
                                 levels = results_summary$sample)

results_summary %>%
    summarise(min=min(V1),
              median=median(V1),
              max=max(V1)) %>%
    gt() %>%
    tab_header(
        title = md("**Summary of variant genes called in B-ALL cohort**")
    ) %>%
    tab_options(
        table.font.size = 12
    ) %>%
    cols_label(
        min = md("**Min**"),
        median = md("**Median**"),
        max = md("**Max**")
    )

ggplot(results_summary, aes(sample, V1)) + 
    geom_bar(position=position_dodge(width=0.8), stat="identity") +
    theme_bw() + xlab("") + ylab("# variant genes") +
    scale_fill_brewer(palette = "Set2") +
    theme(legend.position = "bottom",
          axis.text.x = element_text(size = 7, angle = 90))
```

## ALL-associated genes called

Supplementary Figure 5 in the MINTIE paper. Shows recurrently called variants in
ALL-associated genes in RCH B-ALL cohort. 

```{r ALL_associated_genes}
all_gene_results <- filter(results_by_gene, gene %in% all_genes) %>%
                        collate_vartypes()

paste("We found", 
      all_gene_results$variant_id %>% unique() %>% length(),
      "variants across",
      all_gene_results$gene %>% unique() %>% length(),
      "unique genes") %>%
    print()

# make list of recurrently mutated genes
var_fields <- c("chr1", "pos1", "strand1", "chr2", "pos2", "strand2",
                 "gene", "variant_type", "class", "sample")
all_gene_results <- all_gene_results %>% select(var_fields) %>% distinct()
recurrent_genes <- group_by(all_gene_results, gene) %>%
                        summarise(var_count = length(sample)) %>%
                        filter(var_count > 4) %>%
                        arrange(desc(var_count))

# make summary data frame
all_gene_summary <- group_by(all_gene_results, gene, class, sample) %>%
                        summarise(var_count = length(sample)) %>%
                        filter(gene %in% recurrent_genes$gene)
all_gene_summary$gene <- factor(all_gene_summary$gene,
                                levels = recurrent_genes$gene)

# define category colours and plot
cols <- c("#87649aff",
          "#bdd888ff",
          "#e7d992ff",
          "#bdbdbd")
names(cols) <- c("Fusion",
                 "Transcribed structural variant",
                 "Novel splice variant",
                 "Unknown")
ggplot(all_gene_summary, aes(gene, var_count, fill = class)) +
    geom_bar(sta = "identity") +
    theme_bw() +
    xlab("") +
    ylab("Variants") +
    scale_fill_manual(values = cols) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 90))

# print stats of top 3 gene
all_gene_summary %>%
    group_by(gene) %>%
    summarise(total_vars = sum(var_count)) %>%
    pull(gene) %>%
    as.character() %>%
    head(3) %>%
    lapply(., get_gene_stats, all_gene_summary) %>%
    unlist() %>%
    str_c("\n") %>%
    paste0(collapse = "") %>%
    cat()
```

